The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free eLearning program CrypTool (version 1.4.00)} \vspace{60pt}

### 2007-04-23 script/en/chapters/numbertheory.tex
be changes dated 2006-04-09 2006-04-11 2006-05-05 2006-05-30 2006-07-26

\newpage
\hypertarget{Chapter_ElementaryNT}{}

\section{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003, May 2005) \\

@@

Additionally I added some qualified statements about the security of the RSA algorithm.
\vskip +40 pt

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letters $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letters $i$ ~\mbox{and} $j$ represent natural numbers. 
\item The letters $p$ always represents a prime number.
\item Te sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

%\begin{quote}
%{\em Joanne K. Rowling\footnote{Joanne K. Rowling,~``Harry Potter and the Philosopher's Stone'', Carlsen, (c) 
%1997, chapter ``Through the trapdoor'', p. 310.}:}\newline This isn't magic – it's logic – a puzzle. A 
%lot of the greatest wizards haven't got an ounce of logic.
%\end{quote}

@@

Every whole number has divisors or factors. The number 1 has just one – itself, whereas the number 12 has the 
six factors 1, 2, 3, 4, 6 and 12\footnote{Due to the fact that 12 has so many factors, this number -- and multiples 
of this number -- is often found in everyday life: the 12-hour scale on clocks, the 60 minutes in an hour, the 
360-degree scale for measuring angles, etc. If we divide these scales into segments, the segments often turn out to be whole numbers. 
These are easier to use in mental arithmetic than fractions.}. Many numbers are only divisible by themselves and 
by 1. When it comes to multiplication, these can be regarded as the ``atoms'' in the realm of numbers.

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Chapter_ElementaryNT_5_2}{}
\label{Label_Chapter_ElementaryNT_5_2}

@@

% --------------------------------------------------------------------------
\subsubsection{Addition and multiplication} \index{Addition} \index{Multiplication}\label{addmult}

@@

% --------------------------------------------------------------------------
\subsubsection*{Example of an multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix A of this chapter}). Alternatively, we can also use the primitive roots. }.

@@

% --------------------------------------------------------------------------
\subsubsection{Raising to the power} \index{Raising to the power}

In modular arithmetic, raising to the power is defined as repeated multiplication -- as usual except that 
multiplication is now slightly different.  We can even apply the usual rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b. \nonumber
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\subsubsection{Fast calculation of high powers} \hypertarget{hohpot}{} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See chapter \ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

If $J(m)$ cannot be determined\footnote{According to the first fundamental theorem of number 
theory\index{Number theory!fundamental theorem} and theorem \ref{thm-zth-phinum}, we determine $J(m)$ by reducing $m$ to prime factors\index{Prime factor!decomposition}.}, it is difficult to calculate 
the $c$-th root. This forms the basis for the security assumption used by the RSA encryption system (see 
Sub-section 4.3.1: \hyperlink{RSAVerfahren}{the RSA procedure} and \hyperlink{RSABeweis}{Sub-section 3.10 Proof of the RSA procedure}).
\end{itemize}
The time required for inverting addition and multiplication, on the other hand, 
is simply proportional to $\log m$ or $(\log m)^2$. Powers (for a number $x$ 
calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions 
(See Overview of the one way functions in this \hyperlink{OneWayFunktion1}{Script} and
\hyperlink{OneWayFunktion2}{article}). \index{One way function}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Euler function, Fermat's little theorem and Euler-Fermat}

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Chapter_ElementaryNT_8}{}\label{patternsandstructures}
As mathematicians investigate the structure \index{Structure}
$a *x \equiv b$ mod $m$ (see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}}
 function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler Leonhard}\index{RSA}
\label{Label_KleinerSatzFermat-chap3}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please see chapter \ref{primality_tests}}.

\hypertarget{KleinerSatzFermat-chap3}{}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat, Pierre}
}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\pagebreak
\subsection{Multiplicative order and primitive roots}
\hypertarget{Chapter_ElementaryNT_9}{} \label{MultOrdPrimitveRoot}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Proof of the RSA procedure with Euler-Fermat}
\hypertarget{Chapter_ElementaryNT_10}{}\index{RSA}\label{rsabeweis}

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA''. However, please note that the abbreviation may be
used in other fields to mean something completely different.}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
(\hyperlink{appendix-movies}
{see appendix C: Movies and Literature with Relation to Cryptography}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

% --------------------------------------------------------------------------
\subsubsection{Basic idea of public key cryptography}
\index{Cryptography!public key}
\hypertarget{Chapter_ElementaryNT_10_1}{}
\label{Label_Chapter_ElementaryNT_10_1}

@@

% --------------------------------------------------------------------------
\subsubsection{How the RSA procedure works}\hypertarget{RSA}{}\label{RSA}

The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem \ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source Initiative
\index{Open Source}, that one can claim patents on algorithms. But there is really
no understanding outside the U.S., that one can get a patent for a special case 
(3 factors) of an algorithm (RSA), although the patent for the general case is
almost expired.}$^,$\footnote{If the two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is not true for all $ m<n$ (although
$ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
Example: If $ n=5^2$ then according to theorem \ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$ 
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

@@

% --------------------------------------------------------------------------
\subsubsection{Proof of requirement 1 (invertibility)}\hypertarget{RSAproof}{}\label{RSAproof}

@@

\begin{figure}[!hb]
\begin{center}
\frame{
\includegraphics[scale=0.70, clip, viewport=50 20 700 430]{figures/PrognoseRSAFaktorisierungSecorvo.pdf}
%                                           links unten rechts oben
}
\caption{Forecast about future factorisation records compared with current results
(from Secorvo)}
\label{secorvo-factorisation-forecast}
\end{center}
\end{figure}

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\vspace{4ex}
\begin{center}
\fbox{\parbox{15cm}{
    \emph{Hermann Hesse\footnotemark:}\\
    To let the possible happen, you again and again have to try the impossible.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  Hermann Hesse, German/Swiss writer and Nobel Prize winner, July 2, 1877 $-$ August 9, 1962.}
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

@@

\vskip +10pt
The last records\footnote{%
The 'RSA numbers' are certain large semiprime numbers (i.e., numbers with
exactly two prime factors). They were generated and published by the 
company RSA Security and they form the basis of the RSA Factoring 
Challenge, in which factorisations for these numbers are sought.
See \href{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}
 {\texttt{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}}.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|ccccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & & {\bf Factored on} & {\bf Factored by} \\
\hline
	RSA-155	&	155 & 512 & & August, 1999 & Herman te Riele et al. \\
	C158 & 		158 & 523 & & January, 2002 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & & April, 2003 & Jens Franke et al. \\
	RSA576 & 	174 & 576 & & December, 2003 & Jens Franke et al. \\
	C176 & 		176 & 583 & & May, 2005 & Kazumaro Aoki et al. \\
	RSA-200 & 	200 & 663 & & May, 2005 & Jens Franke et al. \\
\hline
\end{tabular}
\caption{The current general purpose factoring records (as of May 2005)}
\end{center}
\end{table}

@@

% be_2005_UPDATEN_if-new-mersenne-prime-appears

Length in bits of the current world records (RSA-200 $\longleftrightarrow{}$ M-42):
  ~  $663 \longleftrightarrow 25,964,951$.

@@

@@

% --------------------------------------------------------------------------
\subsubsection{One way functions}\index{One way function}
\hypertarget{OneWayFunktion2}{}%
If the key is to be impossible to determine even with knowledge of both the 
plaintext and the cipher text, we need a function that is, on the one hand, 
relatively easy to calculate -- we don't want to have problems encrypting 
messages. On the other hand, the inverse function should exist (otherwise 
information would be lost during encryption), but should be de facto 
incalculable.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{The RSA procedure with actual numbers}\hypertarget{Chapter_ElementaryNT_12}{}
\index{RSA}\label{rsaconcrete}

Having described above \hyperlink{RSA}{how the RSA procedure works}, we will now work through the steps using actual, but small, numbers.

@@

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_A}{} \label{Appendix_A}
\subsection*{Appendix A: the greatest common divisor (gcd) of whole numbers and the two
Algorithms of Euclid} 
\addcontentsline{toc}{subsection}{Appendix A: gcd of whole numbers and the two
algorithms of Euclid}\index{Euclidean algorithm!extended}\index{Gcd}

@@

This appendix describes how to generate arbitrary base representations of anay positive integer and how 
to determine the number of required digits via the logarithm function.

>>>>

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\section{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003, May 2005, March 2006) \\

@@

Additionally I added some qualified statements about the security of the RSA algorithm.

\vskip +40 pt

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letters $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letters $i$ ~\mbox{and} $j$ represent natural numbers. 
\item The letters $p$ always represents a prime number.
\item The sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

%\begin{quote}
%{\em Joanne K. Rowling\footnote{Joanne K. Rowling,~``Harry Potter and the Philosopher's Stone'', Carlsen, (c) 
%1997, chapter ``Through the trapdoor'', p. 310.}:}\newline This isn't magic -- it's logic -- a puzzle. A 
%lot of the greatest wizards haven't got an ounce of logic.
%\end{quote}

@@

Every whole number has divisors or factors. The number 1 has just one -- itself, whereas the number 12 has the 
six factors 1, 2, 3, 4, 6 and 12\footnote{Due to the fact that 12 has so many factors, this number -- and multiples 
of this number -- is often found in everyday life: the 12-hour scale on clocks, the 60 minutes in an hour, the 
360-degree scale for measuring angles, etc. If we divide these scales into segments, the segments often turn out to be whole numbers. 
These are easier to use in mental arithmetic than fractions.}. Many numbers are only divisible by themselves and 
by 1. When it comes to multiplication, these can be regarded as the ``atoms'' in the realm of numbers.

@@

% --------------------------------------------------------------------------
\hypertarget{Chapter_ElementaryNT_5_2}{}
\subsubsection{Patterns and structures}
\label{Label_Chapter_ElementaryNT_5_2}

@@

% --------------------------------------------------------------------------
\subsubsection{Addition and multiplication} \index{Addition} \index{Multiplication}
\label{addmult}

@@

% --------------------------------------------------------------------------
\subsubsection*{Example of a multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix A of this chapter}). Alternatively, we can also use the primitive roots.}.

@@

% --------------------------------------------------------------------------
\subsubsection{Raising to the power} \index{Raising to the power} 
In modular arithmetic, raising to the power is defined as repeated multiplication -- as usual except that 
multiplication is now slightly different.  We can even apply the usual rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b. \nonumber
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsubsection{Fast calculation of high powers} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See chapter \ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

If $J(m)$ cannot be determined\footnote{According to the first fundamental
theorem of number theory\index{Number theory!fundamental theorem} and
theorem \ref{thm-zth-phinum}, we can determine $J(m)$ by reducing $m$ to
prime factors\index{Prime factor!decomposition}.
}, it is difficult to calculate the $c$-th root. This forms the basis for the
security assumption used by the RSA encryption system (see chapter
\ref{rsabeweis} or chapter \ref{rsaverfahren}).

\end{itemize}
The time required for inverting addition and multiplication, on the other hand, 
is simply proportional to $\log m$ or $(\log m)^2$. Powers (for a number $x$ 
calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions 
(See Overview of the one way functions in this \hyperlink{OneWayFunktion1}{Script} and
\hyperlink{OneWayFunktion2}{article}). \index{One way function}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_8}{}
\subsection{Euler function, Fermat's little theorem and Euler-Fermat}

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
\label{L-Euler-Function}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}
} function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler, Leonhard}\index{RSA}
\label{Label_KleinerSatzFermat-chap3}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please see chapter \ref{primality_tests}}.

\hypertarget{KleinerSatzFermat-chap3}{}
\index{Fermat!little theorem}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat, Pierre}
}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\pagebreak
\hypertarget{Chapter_ElementaryNT_9}{}
\subsection{Multiplicative order and primitive roots}
\label{MultOrdPrimitveRoot}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_10}{}
\subsection{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA'' $-$ the following remark is meant with humor to show
that each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
%(\hyperlink{appendix-movies}
%{see appendix A3: Movies and Literature with Relation to Cryptography}).
(see appendix \ref{s:appendix-movies}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

% --------------------------------------------------------------------------
\hypertarget{Chapter_ElementaryNT_10_1}{}
\subsubsection{Basic idea of public key cryptography}
\label{Label_Chapter_ElementaryNT_10_1}
\index{Cryptography!public key}

@@

% --------------------------------------------------------------------------
\hypertarget{RSA}{}
\subsubsection{How the RSA procedure works}
\label{RSA}
The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem \ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source Initiative
\index{Open Source}, that one can claim patents on algorithms. But there is really
no understanding outside the U.S., that one can get a patent for a special case 
(3 factors) of an algorithm (RSA), although the patent for the general case is
almost expired.}$^,$\footnote{If the two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is not true for all $ m<n$ (although
$ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
Example: If $ n=5^2$ then according to theorem \ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$ 
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

@@

% --------------------------------------------------------------------------
\hypertarget{RSAproof}{}
\subsubsection{Proof of requirement 1 (invertibility)}
\label{RSAproof}

@@

\begin{figure}[!hb]
\begin{center}
\frame{
\includegraphics[scale=0.65, clip, viewport=-20 0 680 520]{figures/PrognoseRSAFaktorisierungSecorvo.pdf}
}
\caption{Forecast about future factorisation records compared with current results
(from Secorvo)}
\label{secorvo-factorisation-forecast}
\end{center}
\end{figure}

\clearpage
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\vspace{4ex}
\begin{center}
\fbox{\parbox{15cm}{
    \emph{Hermann Hesse\footnotemark:}\\
    To let the possible happen, you again and again have to try the impossible.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  Hermann Hesse, German/Swiss writer and Nobel Prize winner, July 2, 1877 $-$ August 9, 1962.}
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

@@

\vskip +10pt
The last records\footnote{%
The 'RSA numbers' are certain large semiprime numbers (i.e., numbers with
exactly two prime factors)\index{Number!semi prime}. 
They were generated and published by the 
company RSA Security and they form the basis of the RSA Factoring 
Challenge, in which factorisations for these numbers are sought.
See \href{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}
 {\texttt{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}}.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|ccccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & & {\bf Factored on} & {\bf Factored by} \\
\hline
	RSA-155	&	155 & 512 & & August, 1999 & Herman te Riele et al. \\
	C158 & 		158 & 523 & & January, 2002 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & & April, 2003 & Jens Franke et al. \\
	RSA576 & 	174 & 576 & & December, 2003 & Jens Franke et al. \\
	C176 & 		176 & 583 & & May, 2005 & Kazumaro Aoki et al. \\
	RSA640\footnotemark & 	193 & 600 & & November, 2005 & Jens Franke et al. \\
	RSA-200 & 	200 & 663 & & May, 2005 & Jens Franke et al. \\
\hline
\end{tabular}
\caption{The current general purpose factoring records (as of November 2005)}    % Eyecatcher_neue-Mersenne 
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.\\
RSA Labs offers its challenges since the beginning of the 1990th.
The next bigger challenge awarded with 30,000 US dollars is RSA-704.\\
See:
\href{http://www.heise.de/newsticker/meldung/print/65957}
{\texttt{http://www.heise.de/newsticker/meldung/print/65957}}
}

@@

% be_2005_UPDATEN_if-new-mersenne-prime-appears % Eyecatcher_neue-Mersenne
Length in bits of the current world records:\\
% (RSA-200 $\longleftrightarrow{}$ 43rd known Mersenne prime) \\
%  ~  $663 \longleftrightarrow 30,402,457$.
%$$ (RSA-200 ~~ \longleftrightarrow{} ~~ 43rd ~known~Mersenne~Prime) $$ mnmnmn
%$$ (RSA~-200 ~~ \longleftrightarrow{} ~~ 43rd ~known~Mersenne~Prime) $$
$$ [RSA{-}200~Number] ~~\longleftrightarrow{}~~ [43rd ~known~Mersenne~Prime] $$
$$ 663 ~~ \longleftrightarrow{} ~~ 30,402,457 ~~~~~$$

@@

A detailed explanation from these two authors also can be found in the
RSA Laboratories CryptoBytes \cite{Shamir2003a}.

@@

% --------------------------------------------------------------------------
\hypertarget{OneWayFunktion2}{}%
\subsubsection{One way functions}\index{One way function}
If the key is to be impossible to determine even with knowledge of both the 
plaintext and the cipher text, we need a function that is, on the one hand, 
relatively easy to calculate -- we don't want to have problems encrypting 
messages. On the other hand, the inverse function should exist (otherwise 
information would be lost during encryption), but should be de facto 
incalculable.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_12}{}
\subsection{The RSA procedure with actual numbers}
\label{rsaconcrete}\index{RSA}
Having described above \hyperlink{RSA}{how the RSA procedure works}, we will now work through the steps using actual, but small, numbers.

@@

\bibitem[Shamir2003a]{Shamir2003a} \index{Shamir 2003a} \index{TWIRL device} 
    Adi Shamir, Eran Tromer, \\
    {\em On the Cost of Factoring RSA-1024}, 
    RSA Laboratories CryptoBytes Volume 6, No. 2, Summer 2003, p. 11-20 \\
    \href{http://www.rsasecurity.com/rsalabs/cryptobytes/CryptoBytes\_August\_2003.pdf}
    {\texttt{http://www.rsasecurity.com/rsalabs/cryptobytes/CryptoBytes\_August\_2003.pdf}}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_A}{}
\subsection*{Appendix A: the greatest common divisor (gcd) of whole
             numbers and the two Algorithms of Euclid}
\label{Appendix_A}
\addcontentsline{toc}{subsection}{Appendix A: gcd of whole numbers and the two
algorithms of Euclid}\index{Euclidean algorithm!extended}\index{Gcd}

@@

This appendix describes how to generate arbitrary base representations of
any positive integer and how to determine the number of required digits
via the logarithm function.
